Speeding up decimal multiplication

### 2020-11-21 paper/paper.tex
Update

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{mathrsfs}
\usepackage{cite}
\usepackage[ruled]{algorithm2e}

@@

It is a common theme in practical analysis of algorithms that ``fancy'' algorithms are slow when $n$ is small.
For example, virtually all partical implementations of the ``fancy'' sorting algorithms that are divide-and-conquer in nature --- quick sort and merge sort --- do in fact fall back to ``dumb'' quadratic sorts if the size of the input if less than some threshold.
In some cases, it is not even a matter of ``fancy'' versus ``dumb'' dichotomy, but rather there is a ``hierarchy of fanciness'':
a group of algorithms with different asymptotic behavior such that it makes sense to use each one of them only for some range of values of $n.$
To give an example, \textbf{libgmp} employs the following ``hierarchy of fanciness'' for multiplication:
first goes the ``basecase'' (quadratic) algorithm; then, the Karatsuba algorithm; then, variations of the Toom-Cook algorithm (Toom-3; then Toom-4; then Toom-6.5; then Toom-8.5); finally, the Schönhage--Strassen algorithm (which is Fourier transform-based).

@@

The convolution theorem says that, for any two vectors $x, y \in \mathcal{F}^N,$
\begin{equation} \label{eq:conv_theorem}
x \star y = \mathfrak{f}^{-1}\big( \mathfrak{f}(x) \cdot \mathfrak{f}(y) \big),
\end{equation}
where $\cdot$ denotes scalar (element-wise) product.

@@

Although an $O(N \log N)$ FFT algorithm exists that works for arbitrary $N,$ even prime,
the most widely used FFT algorithm, the Cooley---Tukey algorithm, works by re-writing a transform
of composite length $N = N_1 N_2$ into smaller transforms of lengths $N_1$ and $N_2.$
Thus, it works only for the values of $N$ which are highly composite;
particularly, if $N$ is a power of two, its variants named
decimation in time (DIT) and decimation in frequency (DIF), can be used.

@@

The main drawback of this algorithm is precision issues.
Formally, it is not even a sound algorithm for multiplication: provided that the precision of
your floating-point values is bounded, you can not multiply arbitrarily large numbers with it: at
some point, you are going to get a round-off error large enough to produce a wrong digit in the
answer.

In order to get a provably correct answer, you have to put less bits of information into the
floating-point values than otherwise possible. Because of that, in order to achieve performance
parity with multiplication algorithms that abstain from use of floating point, platform-specific
SIMD extensions are often used. Also, in order to get a provably correct result, you need to know
the precision of your floating-point types, as well as possible quirks of their implementation ---
i.e., depend on the platform.

@@

\subsection{Choosing the base}

@@

The naïve method of modular reduction is to use, where it is available, a hardware instruction
to divide two-words divident by single-word divisor into single-word quotient and single-word
remainder. Where not available, we would have to emulate such an instruction in software.

@@

A method intended to be faster, while not requiring a change in our representation of field
elements, is known as Barrett reduction. % TODO cite

@@

A method even faster for our purposes, is Montgomery reduction.

@@

The \textbf{mpdecimal} library uses yet another method of modular reduction on 64-bit systems;
it is based on the use of prime moduli of form $2^{64} - 2^{n} + 1.$
Generally, such primes are knows as Solinas primes, defined as primes of form
$f(2^m),$ where $f(x)$ is a low-degree polynomial with small integer coefficients.

@@

DFT of length $N = RC$ can be computed as follows:
\begin{enumerate}
    \item $[R \times C]$, perform (length $R$) DFT on each column with $\xi = (\xi_N)^C.$
    \item $[R \times C]$, multiply each element indexed $\langle i,j \rangle$ by $(\xi_N)^{ij}.$
    \item $[R \times C]$, perform (length $C$) DFT on each row with $\xi = (\xi_N)^R.$
    \item $[R \times C]$, transpose the matrix.
\end{enumerate}

@@

DFT of length $N = RC$ can be computed as follows:
\begin{enumerate}
    \item $[R \times C],$ transpose the matrix.
    \item $[C \times R],$ perform (length $R$) DFT on each row with $\xi = (\xi_N)^C.$
    \item $[C \times R],$ transpose the matrix.
    \item $[R \times C],$ multiply each element indexed $\langle i,j \rangle$ by $(\xi_N)^{ij}.$
    \item $[R \times C],$ perform (length $C$) DFT on each row with $\xi = (\xi_N)^R.$
    \item $[R \times C],$ transpose the matrix.
\end{enumerate}

@@

The six-step algorithm has the advantage that, aside from matrix transpositions, it only accesses
memory in strides of $R$ or $C.$ Thus, it might be faster in settings of hierarchical memory,
including external-memory transforms and modern systems with multiple levels of cache.

@@

In its general form, in-place transposition of a non-square matrix is quite hard.
The classical study begins with defining the function $\mathcal{P}(a)$ such that,
when transposing a matrix with $N_1$ rows and $N_2$ columns in row-major order,
the element with index $a$ is sent to index $\mathcal{P}(a).$
There is, then, a result which says that
\begin{equation}
\mathcal{P}(a) = \begin{cases}
    N_1 N_2 - 1                 & \text{if } a = N_1 N_2 - 1; \\
    N_1 a \mymod (N_1 N_2 - 1)  & \text{otherwise.}
\end{cases}
\end{equation}
The number of fixed points of $\mathcal{P}$ is exactly
$$
1 + \gcd(N_1-1, N_2-1).
$$
The number of cycles of length $k>1$ of $\mathcal{P}$ is given by
$$
\frac{1}{k} \sum\limits_{d | k} \widetilde{\mu}(k/d) \gcd(N_1^d - 1, N_1 N_2 - 1),
$$
where $\widetilde{\mu}$ is the Möbius function and the summation is performed over all divisors of
$k.$

Most algorithms for in-place transposition are essentially of ``follow-the-cycles'' kind.
This means that they iterate over all the cycles, and for each cycle,
they shift its elements cyclically. There are, then, various approaches for
locating the cycles, identifying the first element of a cycle, and tracking which cycles were
already visited.
If we limit ourselves to using only $O(N_1+N_2)$ additional memory,
even more complex algorithms are genreally needed;
\cite{Fich95permutingin}
presents an algorithm with worst-case complexity of $O(N_1 N_2 \log(N_1 N_2)).$

@@

Observe that steps 1 and 2 of the direct transform, namely
\begin{itemize}
\item length-3 DFT on each column,
\item multiplication by $(\xi_N)^{ij}$),
\end{itemize}
as well as steps 3 and 4 of the inverse transform, namely
\begin{itemize}
\item multiplication by $(\xi_N)^{-ij},$
\item length-3 IDFT on each column,
\end{itemize}
can be merged (performed in one pass over the columns).
In fact, they should be merged, as it produces less loads/stores and leads to better
cache utilization.

@@

Note that we do need to multiply everything by $N^{-1}$ after performing the inverse transform
(or by another factor if the linearity trick is used; see section~\ref{section:trick}).
For the direct transform, we only need to ``multiply'' (perform $\redc$ with $R^2$)
if the linearity trick is not used.

Suppose we need to perform length-3 DFT on each column, then multiply each matrix element indexed
$\langle i,j \rangle$
by $(\xi_N)^{ij};$ and also multiply everything by a factor $\Phi.$
Define $\Gamma = \xi_N,$ $\Lambda = (\xi_N)^{N/3}.$
This operation on a signle column with index $j,$
$\langle x, y, z \rangle \mapsto \langle x', y', z' \rangle,$ can be expresed as:

\begin{subequations}
\begin{align}
    x' &= \Phi (x + y + z); \\
    y' &= \widetilde{\Phi}_1 (x + y \Lambda   + z \Lambda^2); \\
    z' &= \widetilde{\Phi}_2 (x + y \Lambda^2 + z \Lambda),
\end{align}
\end{subequations}
where
\begin{subequations}
\begin{align}
    \widetilde{\Phi}_1 &= \Phi \Gamma^j; \\
    \widetilde{\Phi}_2 &= \Phi \Gamma^{2j}.
\end{align}
\end{subequations}

@@

Suppose now we need to multiply each matrix element indexed $\langle i,j \rangle$
by $(\xi_N)^{-ij},$ then perform length-3 IDFT on each column; and also multiply everything by a
factor $\Phi.$

Define $\Gamma = (\xi_N)^{-1},$ $\Lambda = (\xi_N)^{-N/3}.$
This operation on a signle column with index $j,$
$\langle x, y, z \rangle \mapsto \langle x', y', z' \rangle,$ can be expresed as:

\begin{subequations}
\begin{align}
    x' &= \widetilde{x} + \widetilde{y}           + \widetilde{z}; \\
    y' &= \widetilde{x} + \widetilde{y} \Lambda   + \widetilde{z} \Lambda^2; \\
    z' &= \widetilde{x} + \widetilde{y} \Lambda^2 + \widetilde{z} \Lambda,
\end{align}
\end{subequations}
where
\begin{subequations}
\begin{align}
    \widetilde{x} &= x \cdot \Phi; \\
    \widetilde{y} &= y \cdot \Phi \Gamma^j; \\
    \widetilde{z} &= z \cdot \Phi \Gamma^{2j}.
\end{align}
\end{subequations}

@@

We need to factorize $N = 2^k$ into $N_1 \cdot N_2 = N$ so that $\max\{N_1, N_2\}$ is as small as
possible.
This is because we will perform transforms of lengths both $N_1$ and $N_2,$ and we want to achieve
good cache utilization.
This is done by choosing $N_1 = 2^{\lfloor k / 2 \rfloor}, N_2 = 2^{\lceil k / 2 \rceil}.$
Note that in case of even $k,$ the matrix is square; in case of odd $k,$
$N_2 = 2 \cdot N_1$ and the matrix can be transposed with the algorithm we described in
section~\ref{section:mat_trans_algo}.

@@

\subsection{Modular addition and subtraction}

@@

By~\eqref{eq:ci_bound}, we have $c_i \le L;$ and by~\eqref{eq:sigmai_bound}, we have
$\sigma_i < (B-1)M \le L.$ But then
$$\sigma_i + c_i < 2L < \mu^2 / 2.$$

Note this is far from being a tight bound in numerical sense:
in practice, the value of
$\frac{L}{(B-1)M}=B-1$ is going to be much larger than 1, meaning that the exact upper bound on
$c_i + \sigma_i$ is going to be closer to $\mu^2 / 4$ than to $\mu^2 / 2;$ but this is still a
tight bound in the sense that the maximum possible $c_i + \sigma_i$ would overflow one machine word.
This bound also has the advantage that it works for any $B > 1.$

>>>>

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{mathrsfs}
\usepackage{cite}
\usepackage{url}
\usepackage[ruled]{algorithm2e}

@@

It is a common theme in practical analysis of algorithms that ``fancy'' algorithms are slow when $n$ is small.
For example, virtually all partical implementations of the ``fancy'' sorting algorithms that are divide-and-conquer in nature --- quick sort and merge sort --- do in fact fall back to ``dumb'' quadratic sorts if the size of the input if less than some threshold.
In some cases, it is not even a matter of ``fancy'' versus ``dumb'' dichotomy, but rather there is a ``hierarchy of fanciness'':
a group of algorithms with different asymptotic behavior such that it makes sense to use each one of them only for some range of values of $n.$
To give an example, \textbf{libgmp} employs the following ``hierarchy of fanciness'' for multiplication~\cite{gmp_fh}:
first goes the ``basecase'' (quadratic) algorithm; then, the Karatsuba algorithm; then, variations of the Toom-Cook algorithm (Toom-3; then Toom-4; then Toom-6.5; then Toom-8.5); finally, the Schönhage--Strassen algorithm (which is Fourier transform-based).

@@

The convolution theorem~\cite{conv_theorem} says that, for any two vectors $x, y \in \mathcal{F}^N,$
\begin{equation} \label{eq:conv_theorem}
x \star y = \mathfrak{f}^{-1}\big( \mathfrak{f}(x) \cdot \mathfrak{f}(y) \big),
\end{equation}
where $\cdot$ denotes scalar (element-wise) product.

@@

Although an $O(N \log N)$ FFT algorithm exists that works for arbitrary $N,$ even
prime~\cite{bluestein},
the most widely used FFT algorithm, the Cooley---Tukey algorithm, works by re-writing a transform
of composite length $N = N_1 N_2$ into smaller transforms of lengths $N_1$ and $N_2.$
Thus, it works only for the values of $N$ which are highly composite;
particularly, if $N$ is a power of two, its variants named
decimation in time (DIT) and decimation in frequency (DIF), can be used.

@@

The main drawback of this algorithm is precision issues.
Formally, it is not even a sound algorithm for multiplication: provided that the precision of
your floating-point values is bounded, you can not multiply arbitrarily large numbers with it: at
some point, you are going to get a round-off error large enough to produce a wrong digit in the
answer~\cite{precision_paper}.

In order to get a provably correct answer, you have to put less bits of information into the
floating-point values than otherwise possible~\cite{gmp_paper, mattcomp_precision}.
Because of that, in order to achieve performance
parity with multiplication algorithms that abstain from use of floating point, platform-specific
SIMD extensions are often used~\cite{gmp_paper}.
Also, in order to get a provably correct result, you need to know the precision of your
floating-point types, as well as possible quirks of their implementation --- i.e., depend on the
platform.

@@

\subsection{Choosing the base}
\label{section:bases}

@@

The naïve method of modular reduction is to use, where it is available, a hardware instruction
to divide two-words dividend by single-word divisor into single-word quotient and single-word
remainder. Where not available, we would have to emulate such an instruction in software.

@@

A method intended to be faster, while not requiring a change in our representation of field
elements, is known as Barrett reduction~\cite{barrett}.

@@

A method even faster for our purposes, is Montgomery reduction~\cite{Montgomery1985ModularMW}.

@@

The \textbf{mpdecimal} library uses yet another method of modular reduction on 64-bit systems;
it is based on the use of prime moduli of form $2^{64} - 2^{n} + 1.$
Generally, such primes are knows as Solinas primes, or generalized Mersenne
primes~\cite{Solinas},
defined as primes of form $f(2^m),$ where $f(x)$ is a low-degree polynomial with small integer
coefficients.

@@

DFT of length $N = RC$ can be computed as follows~\cite{mattcomp_mfa}:
\begin{enumerate}
    \item $[R \times C]$, perform (length $R$) DFT on each column with $\xi = (\xi_N)^C.$
    \item $[R \times C]$, multiply each element indexed $\langle i,j \rangle$ by $(\xi_N)^{ij}.$
    \item $[R \times C]$, perform (length $C$) DFT on each row with $\xi = (\xi_N)^R.$
    \item $[R \times C]$, transpose the matrix.
\end{enumerate}

@@

DFT of length $N = RC$ can be computed as follows~\cite{mattcomp_mfa}:
\begin{enumerate}
    \item $[R \times C],$ transpose the matrix.
    \item $[C \times R],$ perform (length $R$) DFT on each row with $\xi = (\xi_N)^C.$
    \item $[C \times R],$ transpose the matrix.
    \item $[R \times C],$ multiply each element indexed $\langle i,j \rangle$ by $(\xi_N)^{ij}.$
    \item $[R \times C],$ perform (length $C$) DFT on each row with $\xi = (\xi_N)^R.$
    \item $[R \times C],$ transpose the matrix.
\end{enumerate}

@@

The six-step algorithm has the advantage that, aside from matrix transpositions, it only accesses
memory in strides of $R$ and $C.$ Thus, it might be faster in settings of hierarchical memory,
including external-memory transforms and modern systems with multiple levels of cache.

@@

In its general form, in-place transposition of a non-square matrix is quite hard.
The classical study begins with defining the function $\mathcal{P}(a)$ such that,
when transposing a matrix with $N_1$ rows and $N_2$ columns in row-major order,
the element with index $a$ is sent to index $\mathcal{P}(a).$
It can be defined explicitly as follows:
\begin{equation}
\mathcal{P}(a) = \begin{cases}
    N_1 N_2 - 1                 & \text{if } a = N_1 N_2 - 1; \\
    N_1 a \mymod (N_1 N_2 - 1)  & \text{otherwise.}
\end{cases}
\end{equation}
There is, then, a result~\cite{fancything} saying that
the number of fixed points of $\mathcal{P}$ is exactly
$$
1 + \gcd(N_1-1, N_2-1);
$$
and the number of cycles of length $k>1$ of $\mathcal{P}$ is
$$
\frac{1}{k} \sum\limits_{d | k} \widetilde{\mu}(k/d) \gcd(N_1^d - 1, N_1 N_2 - 1),
$$
where $\widetilde{\mu}$ is the Möbius function and the summation is performed over all divisors of
$k.$

Most algorithms for in-place transposition are essentially of ``follow-the-cycles'' kind.
This means that they iterate over all the cycles, and for each cycle,
they shift its elements cyclically. There are, then, various approaches for
locating the cycles, identifying the first element of a cycle, and tracking which cycles were
already visited.
If we limit ourselves to using only $O(N_1+N_2)$ additional memory,
even more complex algorithms are generally needed;
\cite{Fich95permutingin}
presents an algorithm with worst-case complexity of $O(N_1 N_2 \log(N_1 N_2)).$

@@

Observe that steps 1 and 2 of the direct transform:
\begin{itemize}
\item length-3 DFT on each column,
\item multiplication by $(\xi_N)^{ij},$
\end{itemize}
as well as steps 3 and 4 of the inverse transform:
\begin{itemize}
\item multiplication by $(\xi_N)^{-ij},$
\item length-3 IDFT on each column,
\end{itemize}
can be merged --- performed in a single pass over the columns.
In fact, they \textit{should} be merged, as it produces less loads/stores and leads to better
cache utilization.

@@

Note that we are going to need to multiply everything by $N^{-1}$ after performing the inverse
transform (or by another factor if the linearity trick is used; see section~\ref{section:trick}).
For the direct transform, we only need to ``multiply'' (perform $\redc$ with $R^2$)
if the linearity trick is not used.

First, consider the case of direct transform.
Suppose we need to perform length-3 DFT on each column, then multiply each matrix element indexed
$\langle i,j \rangle$
by $(\xi_N)^{ij};$ and also multiply everything by a factor $\Phi.$

Define $\Gamma = \xi_N,$ $\Lambda = (\xi_N)^{N/3}.$
This operation on a single column with index $j,$
$\langle x, y, z \rangle \mapsto \langle x', y', z' \rangle,$ can be expressed as:
\begin{subequations}
\begin{align}
    x' &= \Phi (x + y + z); \\
    y' &= \widetilde{\Phi}_1 (x + y \Lambda   + z \Lambda^2); \\
    z' &= \widetilde{\Phi}_2 (x + y \Lambda^2 + z \Lambda),
\end{align}
\end{subequations}
where
\begin{subequations}
\begin{align}
    \widetilde{\Phi}_1 &= \Phi \Gamma^j; \\
    \widetilde{\Phi}_2 &= \Phi \Gamma^{2j}.
\end{align}
\end{subequations}

@@

Now, the case of inverse transform.
We now need to multiply each matrix element indexed $\langle i,j \rangle$
by $(\xi_N)^{-ij},$ then perform length-3 IDFT on each column; and also multiply everything by a
factor $\Phi.$

Define $\Gamma = (\xi_N)^{-1},$ $\Lambda = (\xi_N)^{-N/3}.$
This operation on a single column with index $j,$
$\langle x, y, z \rangle \mapsto \langle x', y', z' \rangle,$ can be expressed as:
\begin{subequations}
\begin{align}
    x' &= \widetilde{x} + \widetilde{y}           + \widetilde{z}; \\
    y' &= \widetilde{x} + \widetilde{y} \Lambda   + \widetilde{z} \Lambda^2; \\
    z' &= \widetilde{x} + \widetilde{y} \Lambda^2 + \widetilde{z} \Lambda,
\end{align}
\end{subequations}
where
\begin{subequations}
\begin{align}
    \widetilde{x} &= x \cdot \Phi; \\
    \widetilde{y} &= y \cdot \Phi \Gamma^j; \\
    \widetilde{z} &= z \cdot \Phi \Gamma^{2j}.
\end{align}
\end{subequations}

@@

Remember that the reason we might turn to using the six-step algorithm is its cache
friendliness: if we interpret the sequence as $N_1 \times N_2$ matrix, then, aside from matrix
transpositions, it only accesses memory in strides of $N_1$ and $N_2.$

With $N = 2^k,$
we need to factorize $N$ into $N_1 N_2 = N$ so that $N_1$ and $N_2$ are as close to each other
as possible.

This can be done by choosing $N_1 = 2^{\lfloor k / 2 \rfloor}, N_2 = 2^{\lceil k / 2 \rceil}.$
Note that in case of even $k,$ the matrix is square; in case of odd $k,$
$N_2 = 2 \cdot N_1$ and the matrix can be transposed with the algorithm we described in
section~\ref{section:mat_trans_algo}.

@@

\subsection{Modular addition, subtraction, and adjustment}

For the purposes of this subsection, ``adjustment'' means reducing a value in range $[0; 2p-1]$
modulo $p$ to be in range $[0; p-1].$

Using actual conditional jumps on assembly level for modular addition, subtraction, or adjustment,
is slow on most modern architectures:
modern processors employ branch prediction, and branches of this
sort are anything but predictable. This results in lots of costly branch mispredictions.

We describe two possible ways to resolve this problem.

\subsubsection{Bit wizardry}

We assume here that two's complement is used,
all values are $w$-bit,
and a modulo $p < 2^w / 2$ is fixed.

For a signed $w$-bit value $x,$ we define $\underline{\mathrm{ExtractSign}}(x)$ as follows:
$$
\underline{\mathrm{ExtractSign}}(x) =
\begin{cases}
0, & \text{if } x \ge 0; \\
-1, & \text{if } x < 0.
\end{cases}
$$
We can compute $\underline{\mathrm{ExtractSign}}(x)$ without branches as follows:
perform signed right shift of $x$ by $(w-1)$ bits.
This leads to filling all $w$ bits with the sign bit of $x;$
all zeros result in $0,$ while all ones result in $-1.$
We can then use this value as a mask.

For $x \in [-p; p-1],$ we define
$\underline{\mathrm{AdjustSigned}}(x)$ as follows:
$$
\underline{\mathrm{AdjustSigned}}(x) =
\begin{cases}
x, & \text{if } x \ge 0; \\
x+p, & \text{if } x < 0.
\end{cases}
$$

It can be computed without branches as follows:
$$\underline{\mathrm{AdjustSigned}}(x) = x + (p \mathbin{\&} \underline{\mathrm{ExtractSign}}(x)),$$
where $\mathbin{\&}$ denotes bitwise ``AND''.
Note that $p \mathbin{\&} 0 = 0,$ $p \mathbin{\&} -1 = p.$

The modular addition of $a,b \in [0; p-1]$ can then be expressed as
$$\underline{\mathrm{AdjustSigned}}(a+b-p).$$
The normal, ``unsigned'' adjustment of $a \in [0; 2p-1]$ is expressed as
$$\underline{\mathrm{AdjustSigned}}(a-p).$$
The modular subtraction of $a,b \in [0; p-1]$ is expressed as
$$\underline{\mathrm{AdjustSigned}}(a-b).$$

\subsubsection{Conditional moves}

The x86 and x86-64 platforms provide ``conditional moves'' --- instructions that
store a value in a register conditionally, depending on a flag.
They do not alter control flow and are not subject to branch prediction.
Various versions of ARM have similar capabilities, although differently named.

Since version 3.8 (latest 11.0), the Clang compiler has built-in called
\verb~__builtin_unpredictable(x)~. According to the documentation,
it ``is used to indicate that a branch condition is unpredictable by hardware mechanisms such as
branch prediction logic''. We used it to implement modular addition, subtraction, and adjustment,
which resulted in 8---13\% speedup of the overall multiplication procedure.

Curiously, \verb~__builtin_unpredictable~ does not seem to produce any effect at all:
replacing \verb~__builtin_unpredictable(x)~ with \verb~(x)~ does not affect the generated code
at all. This means that Clang is smart enough to replace the branch with a conditional move
even without our hint. But this does not mean, however, that we should just write the code with
the branch and rely on the compiler to figure it out; compiling the code with branches with GNU
GCC 10.2.0 resulted in code 40---60\% slower than the one compiled with Clang, whilst with
the ``bit wizardry'' approach, GNU GCC outputs code with performance comparable to Clang.
This is because GNU GCC does not rewrite the branch into a conditional move, assuming, instead,
that it can be predicted by the hardware.

The above should mean, if anything, that compiler optimizations and heuristics are very flacky,
and should never be relied upon.

@@

By~\eqref{eq:ci_bound}, we have $c_i \le L;$ and by~\eqref{eq:sigmai_bound}, we have
$\sigma_i < (B-1)M \le L.$ But then
\begin{equation} \label{eq:two_word_bound}
\sigma_i + c_i < 2L < \mu^2 / 2.
\end{equation}
We see that it fits into two machine words.

Note this is far from being a tight bound in numerical sense:
in practice, the value of
$\frac{L}{(B-1)M}=B-1$ is going to be much larger than 1, meaning that the exact upper bound on
$\sigma_i + c_i$ is going to be closer to $\mu^2 / 4$ than to $\mu^2 / 2.$
But this is still a tight bound in the sense that the maximum possible $\sigma_i + c_i$ would
not fit into a single machine word.
This bound also has the advantage that it works for any $B > 1.$

\subsubsection{Division by constant}

The iterative process~\eqref{eq:sigmai_and_zi} requires calculating both quotient and remainder
of integer division.
By~\eqref{eq:two_word_bound}, the dividend generally occupies two machine words.
The divisor is $B,$ which, for our choice of bases (see section~\ref{section:bases}),
always fits into one machine word.

It is possible the built-in two-word types: \verb~uint64_t~ on 32-bit systems and
\verb~unsigned __int128~ on 64-bit systems. But the latest versions of both Clang and GNU GCC do no
optimize the division of two-word types by a constant, as they do for one-word types. Thus,
this built-in division is slow.

We use the approach described in~\cite{div_by_const} of rewriting a division by a constant
into a sequence of cheaper operations, generally a single double-width multiplication,
some bit shifts, and, for some divisors, addition and subtraction.
For each possible value of $B$ (see, again, section~\ref{section:bases}: we support 3 different
bases on 32-bit systems and 4 different bases on 64-bit systems), we generate the code for answer
recovery that performs division by $B.$

\subsubsection{Chinese remainder theorem}

The explicit solution of finding the remainder $m = ( x \mymod (p_1 \cdot p_2) ),$
where
\begin{subequations}
\begin{align}
x \mymod p_1 &= a_1; \\
x \mymod p_2 &= a_2,
\end{align}
\end{subequations}
is given by
\begin{equation}
m = a_1 + p_1 \cdot \Big( \big( r \cdot (a_2 - a_1) \big) \, \mymod \, p_2 \Big),
\end{equation}
where
$$r \equiv p_1^{-1} \, (\mymod \, p_2).$$

If we sort the primes so that $p_1 < p_2,$
then $( (a_2 - a_1) \mymod p_2 )$
is just a subtraction modulo $p_2.$
We define $\psi = (a_2 - a_1) \mymod p_2.$

Note that we can pre-calculate the value of $r.$ We can also pre-calculate
the Montgomery representation (see section~\ref{section:montgomery_intro})
of $r$ for the second modulo,
$$
\widetilde{r} = (R \cdot r) \mymod p_2.
$$
Then, the value of $( (r \cdot \psi) \mymod p_2 )$ can be calculated as
$$\varphi = \redc(\widetilde{r}, \psi),$$
where the $\redc$ is done for the modulo $p_2.$

The value of $m = a_1 + p_1 \cdot \varphi$ is then computed as usual.
